### 2019-10-22 Manuscript/Manuscript.tex

The SE Asian islands (SEAI) have an outsized contribution on modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic outgassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a}. Here we use geologic data to reconstruct the shorelines of SEAI over the past 15~m.y. (Fig. \ref{fig:shoreline_growth}) and quantify the increase in land area. We use these paleoshorelines as boundary conditions within a coupled weathering-climate model to quantify the change in steady-state \pCOtwo associated with this emergence (Fig. \ref{fig:scenario_pCO2}). We find that the modeled decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain long-term climatic cooling over the Neogene.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in SEAI and enhanced global weatherability, we use the spatially-resolved GEOCLIM model \cite{Godderis2014a, Godderis2017c}. GEOCLIM consists of a silicate weathering model linked to climate model runs at different \pCOtwo levels, allowing for estimates of long-term steady-state \pCOtwo associated with coupled changes in weathering sinks and climatology.

@@

The values of four parameters within the silicate weathering component \cite{Gabet2009a, West2012a, Maffre2018a} that modify the dependence of silicate weathering on temperature, runoff, and particle age are poorly constrained (see SI). Rather than prescribing values for these parameters, we allow them and metamorphic and siliciclastic Ca+Mg concentrations to vary within reasonable bounds, resulting in 78,000 unique combinations. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 3,339 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo outgassing flux \cite{Gerlach2011a} as they should with the long-term carbon cycle being at steady-state (see SI).

@@

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of islands in the area at ca. 15, 10, and 5~Ma (``reduced SEAI + Sunda Shelf'' scenarios; Fig. \ref{fig:scenario_pCO2}). Given that the Sunda Shelf is presently submerged, across the shelf we distribute felsic and siliciclastic lithologies and assign shallow slope values in a manner that is consistent with existing observations, and extrapolate runoff from surrounding lands (see SI). While we modify the land area, the position of the tectonic blocks remain fixed since we use a climate model forced with modern geography. While there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (see SI). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

Using the 3,339 unique parameter combinations, the ``reduced SEAI + Sunda Shelf'' scenarios resulted in 533--611~ppm, 466--502~ppm, and 394--412~ppm \pCOtwo for 15, 10, and 5~Ma respectively (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$533--611~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a, Mejia2017a, Ji2018a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (see SI), indicating that the increase in subaerially-exposed land area and tectonic topography of SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 3,339 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched \COtwo consumption estimated from watershed data (see SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to 1.5$\times$ the interquartile range, or to the minimum/maximum of the estimates if no estimates are below/above this range.}
    \label{fig:scenario_pCO2}
\end{figure}

>>>>

The SE Asian islands (SEAI) have an out-sized contribution on modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic outgassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a}. Here we use geologic data to reconstruct the shorelines of SEAI over the past 15~m.y. (Fig. \ref{fig:shoreline_growth}) and quantify the increase in land area. We use these paleoshorelines as boundary conditions within a coupled weathering-climate model to quantify the change in steady-state \pCOtwo associated with this emergence (Fig. \ref{fig:scenario_pCO2}). We find that the modeled decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain long-term climatic cooling over the Neogene.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area of SEAI and enhanced global weatherability, we use the spatially-resolved GEOCLIM model \cite{Godderis2014a, Godderis2017c}. GEOCLIM consists of a silicate weathering model linked to climate model runs at different \pCOtwo levels, allowing for estimates of long-term steady-state \pCOtwo associated with coupled changes in weathering sinks and climatology.

@@

The values of four parameters within the silicate weathering component \cite{Gabet2009a, West2012a, Maffre2018a} that modify the dependence of silicate weathering on temperature, runoff, and particle age are poorly constrained (see SI). Rather than prescribing values for these parameters, we allow them and metamorphic and siliciclastic Ca+Mg concentrations to vary within reasonable bounds, resulting in 93,600 unique combinations. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 5,381 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo outgassing flux \cite{Gerlach2011a} as they should with the long-term carbon cycle being at steady-state (see SI).

@@

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of islands in the area at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Given that the Sunda Shelf is presently submerged, across the shelf we distribute felsic and siliciclastic lithologies and assign shallow slope values in a manner that is consistent with existing observations, and extrapolate runoff from surrounding land (see SI). While we modify the land area, the position of the tectonic blocks remain fixed since we use a climate model forced with modern geography. While there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (see SI). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

Using the 5,381 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 493--709~ppm, 435--532~ppm, and 378--447~ppm \pCOtwo for 15, 10, and 5~Ma respectively (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$493--709~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a, Mejia2017a, Ji2018a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (see SI), indicating that the increase in subaerially-exposed land area and tectonic topography of SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 5,381 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched \COtwo consumption estimated from watershed data (see SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values.}
    \label{fig:scenario_pCO2}
\end{figure}
